---
title: "Cultural distances: controlling for history"
output: pdf_document
---

```{r echo=F, eval=F}
setwd("~/Documents/Bristol/word2vec/word2vec_DPLACE/analysis/")
```

# Introduction

We compare cultural distances between socieites with linguistic distances between societies, controlling for shared history in two ways.

The first test uses mixed effects modelling.  The pairing of the langauge family of each language (according to Glottolog) is used as a random effect.  That means that the model can capture the likelihood that two languages from the Indo-European langauge family will be more similar to each other than two languages from different langauge families.  The same is done with geographic area according to Autotyp.

The second test controls for history using distances from a phylogenetic tree.  The tree comes from Bouckaert et al. (2012).  Patristic distances between languages are used as a measure of historical distance between societies in a Mantel test.  Note that the Mantel test assumes a strict distance metric, which is not necessarily the case with this data, but there are few other ways to deal with continuous pairwise distances. 

# Load libraries
```{r warning=F, message=F}
library(ape)
library(ecodist)
library(lme4)
library(sjPlot)
library(ggplot2)
library(igraph)
library(lattice)
```

# All domains

## Load data

Read the cultural distances:

```{r}
cult = read.csv("../results/EA_distances/CulturalDistances_Long.csv", stringsAsFactors = F)
names(cult) = c("l1","l2","cult.dist")
cultLangs = unique(c(cult$Var1,cult$Var2))
```

Add language family:

```{r}
l = read.csv("../data/FAIR_langauges_glotto_xdid.csv", stringsAsFactors = F)
g = read.csv("../data/glottolog-languoid.csv/languoid.csv", stringsAsFactors = F)
l$family = g[match(l$glotto,g$id),]$family_pk
l$family = g[match(l$family,g$pk),]$name
```

Read the semantic distances

```{r}
ling = read.csv("../data/FAIR/semantic_distances_FAIR.csv", stringsAsFactors = F)
```

Combine the lingusitic and cultural distances

```{r}
cult$l1.iso2 = l[match(cult$l1,l$Language2),]$iso2
cult$l2.iso2 = l[match(cult$l2,l$Language2),]$iso2

fairisos = unique(c(ling$l1,ling$l2))
cultisos = unique(c(cult$l1.iso2, cult$l2.iso2))

cult = cult[(cult$l1.iso2 %in% fairisos) & (cult$l2.iso2 %in% fairisos),]
ling = ling[(ling$l1 %in% cultisos) & (ling$l2 %in% cultisos),]

matches = sapply(1:nrow(ling), function(i){
  which(cult$l1.iso2==ling$l1[i] & cult$l2.iso2==ling$l2[i])
})

ling$cult.dist = cult[matches,]$cult.dist
ling$cult.dist.center = scale(ling$cult.dist)
cdc.s = attr(ling$cult.dist.center,"scaled:scale")
cdc.c = attr(ling$cult.dist.center,"scaled:center")
ling$cult.dist.center = as.numeric(ling$cult.dist.center)

ling$family1 = l[match(ling$l1, l$iso2),]$family
ling$area1 = l[match(ling$l1, l$iso2),]$autotyp.area

ling$family2 = l[match(ling$l2, l$iso2),]$family
ling$area2 = l[match(ling$l2, l$iso2),]$autotyp.area

ling$family.group = paste(ling$family1,ling$family2)
ling$area.group = paste(ling$area1,ling$area2)

ling$rho.center = scale(ling$rho)

```

Each observation is now assocaited with a langauge family pair:

```{r}
head(ling[,c("l1","l2","rho",'family.group')])
```

And the same is true for area:

```{r}
tail(ling[,c("l1","l2","rho",'area.group')])
```

## LMER models

Mixed effects model, predicting linguistic distances from cultural distances, with random intercept for family and area and random slope for cultural distance for family and area.

We compare a null model to a model with a fixed effect for cultural distance, with random intercepts for family and area, and random slopes for cultural distance by both.

```{r}
m0 = lmer(
  rho.center ~ 1 +
    (1 + cult.dist.center | family.group) +
    (1 + cult.dist.center | area.group),
  data = ling
)
m1 = lmer(
  rho.center ~ 1 +
    cult.dist.center +
    (1 + cult.dist.center | family.group) +
    (1 + cult.dist.center | area.group),
  data = ling
)
anova(m0,m1)
```

Cultural distance is significantly correlated with linguistic distance.  Here are the model estimates:

```{r}
summary(m1)
```

Plot the estimates, rescaling the variables back to the original units:

```{r}
gx = sjp.lmer(m1,'pred','cult.dist.center', prnt.plot = F)
gx$plot$data$y = gx$plot$data$y *
  attr(ling$rho.center,"scaled:scale") +
  attr(ling$rho.center,"scaled:center")
gx$plot$data$resp.y = gx$plot$data$resp.y *
  attr(ling$rho.center,"scaled:scale") +
  attr(ling$rho.center,"scaled:center")
gx$plot$data$x = gx$plot$data$x *
  cdc.s +cdc.c
gx$plot + coord_cartesian(ylim=c(0.1,0.6),
                          xlim=c(0.2,0.8))
```

Plot the random effects.  There seems to be more variation by area:

```{r}
sjp.lmer(m1,'re', sort.est = "cult.dist.center")
px = sjp.lmer(m1,'rs.ri', prnt.plot = F)
dx = px$plot[[1]]$data
dx$x = dx$x * cdc.s + cdc.c
dx$y = dx$y * 
  attr(ling$rho.center,"scaled:scale") +
  attr(ling$rho.center,"scaled:center")
ggplot(dx,aes(x,y)) +
  geom_point(data=ling, 
             mapping=aes(x=as.numeric(cult.dist),
                         y=as.numeric(rho))) +
  geom_line(mapping = aes(colour=grp)) +
  xlab("Cultural distance")+
  ylab("Linguistic distance") +
  ggtitle("Area pair random effects") +
  coord_cartesian(ylim=c(0.1,0.6),
                  xlim=c(0.2,0.8))

dx = px$plot[[2]]$data
dx$x = dx$x * cdc.s + cdc.c
dx$y = dx$y * 
  attr(ling$rho.center,"scaled:scale") +
  attr(ling$rho.center,"scaled:center")
ggplot(dx,aes(x,y)) +
  geom_point(data=ling, 
             mapping=aes(x=as.numeric(cult.dist),
                         y=as.numeric(rho))) +
  geom_line(mapping = aes(colour=grp)) +
  xlab("Cultural distance")+
  ylab("Linguistic distance") +
  ggtitle("Family pair random effects") +
  coord_cartesian(ylim=c(0.1,0.6),
                  xlim=c(0.2,0.8))
```


\newpage

# Tests within domains

Load distances for specific domains and match up to language family and area:

```{r}
ling.dom = read.csv(
  "../results/EA_distances/All_Domains_with_ling.csv", 
  stringsAsFactors = F)

ling.dom = ling.dom[!is.na(ling.dom$cult.dist),]

ling.dom$family1 = l[match(ling.dom$l1, l$iso2),]$family
ling.dom$family2 = l[match(ling.dom$l2, l$iso2),]$family
ling.dom$family.group = paste(ling.dom$family1,
                              ling.dom$family2)

ling.dom$area1 = l[match(ling.dom$l1, l$iso2),]$autotyp.area
ling.dom$area2 = l[match(ling.dom$l2, l$iso2),]$autotyp.area
ling.dom$area.group = paste(ling.dom$area1,
                              ling.dom$area2)
```

Center the data:

```{r}
ling.dom$cult.dist.center = scale(ling.dom$cult.dist)
ling.dom$rho.center = scale(ling.dom$rho)
```

## LMER models

Test whether random slopes are warraneted for family:

```{r}
mD0 = lmer(
  rho.center ~ 1 +
    (1 | family.group) +
    (1 | area.group) +
    (1 | imputed_semantic_domain),
  data = ling.dom)
mD1 = lmer(
  rho.center ~ 1 +
    (1 | family.group) +
    (0 + cult.dist.center | family.group) +
    (1 | area.group) +
    (1 | imputed_semantic_domain),
  data = ling.dom)
mD2 = lmer(
  rho.center ~ 1 +
    (1 + cult.dist.center | family.group) +
    (1 | area.group) +
    (1 | imputed_semantic_domain),
  data = ling.dom)
anova(mD0,mD1,mD2)
```

Random slopes for family improves the fit of the model, but adding the correlation coefficient does not.

Test the same for area:

```{r}
mD3 = lmer(
  rho.center ~ 1 +
    (1 | family.group) +
    (0 + cult.dist.center | family.group) +
    (1 | area.group) +
    (0 + cult.dist.center | area.group) +
    (1 | imputed_semantic_domain),
  data = ling.dom)
anova(mD1,mD3)
```

Random slopes for area do not improve the fit of the model.

```{r}
mD4 = lmer(
  rho.center ~ 1 +
    (1 + cult.dist.center | family.group) +
    (1 | area.group) +
    (1 | imputed_semantic_domain),
  data = ling.dom)
```

And, in fact, area random intercepts are all set to 0:
```{r}
ranef(mD4)$area.group
```

So we'll take them out.

Now we test the main effect of cultural distance:

```{r}
mD5 = lmer(
  rho.center ~ 1 +
    (1 + cult.dist.center || family.group) +
    (1 + cult.dist.center | imputed_semantic_domain),
  data = ling.dom)
mD6 = update(mD5, ~.+cult.dist.center)
anova(mD5,mD6)
```

Summary of the final model, with random effects plot:

```{r}
summary(mD6)
dotplot(ranef(mD6))
```

\newpage

Plot the predicted relationships for each domain.  The domains in the legend are sorted by the slope for cultural distance (greatest negative slope to greatest positive slope):

```{r}
sjp.lmer(mD6,'eff', show.ci = T)
dom.order = ranef(mD6)$imputed_semantic_domain
dom.order = rownames(dom.order[order(dom.order$cult.dist.center),])
px = sjp.lmer(mD6,'rs.ri', show.ci = T, prnt.plot = F)
pdx = px$plot[[2]]$data
pdx$Domain = factor(pdx$grp, levels = dom.order)
pdx$x = pdx$x *
  attr(ling.dom$cult.dist.center,"scaled:scale") +
  attr(ling.dom$cult.dist.center,"scaled:center")
pdx$y = pdx$y *
  attr(ling.dom$rho.center,"scaled:scale") +
  attr(ling.dom$rho.center,"scaled:center")
ggplot(pdx,
       aes(x,y,colour=Domain)) +
  geom_line() +
  xlab("Cultural Distance") +
  ylab("Linguistic distance (rho)")

```

"Religion and beleif" seems to be working differently from the other domains.

\newpage

# Mantel tests

Read the historical distances for Indo-European, based on the phylogenetic distances.

## Data prep

Load historical distances:

```{r}
hist = read.csv("../data/trees/IndoEuropean_historical_distances.csv", stringsAsFactors = F)
hist = hist[!duplicated(hist[,1]),!duplicated(hist[,1])]
rownames(hist) = hist[,1]
hist = hist[,2:ncol(hist)]
hist.m = as.matrix(hist)
colnames(hist.m) = rownames(hist.m)
hist.m = hist.m/max(hist.m)
```

Read the cultural distances as a matrix:

```{r}
cult.m = read.csv("../results/EA_distances/CulturalDistances.csv", stringsAsFactors = F)
rownames(cult.m) = cult.m[,1]
cult.m = cult.m[,2:ncol(cult.m)]
```

Convert the linguistic distances to a matrix.  This uses `igraph` to make an undirected graph from the long format with `rho` as the edge weights, then output a matrix of adjacencies.

```{r}
grph <- graph.data.frame(ling[,c("l1",'l2','rho')], directed=FALSE)
# add value as a weight attribute
ling.m = get.adjacency(grph, attr="rho", sparse=FALSE)
rownames(ling.m) = l[match(rownames(ling.m),l$iso2),]$Language2
colnames(ling.m) = l[match(colnames(ling.m),l$iso2),]$Language2
```

Match the distance matrices

```{r}
in.analysis = intersect(rownames(ling.m),rownames(cult.m))
in.analysis = intersect(in.analysis, rownames(hist.m))
cult.m2 = cult.m[in.analysis,in.analysis]
ling.m2 = ling.m[in.analysis,in.analysis]
hist.m2 = hist.m[in.analysis,in.analysis]
```

Note that there are only `r length(in.analysis)` languages with data on lingusitic, cultural and historical distance.

## Tests

Simple correlation without control for history:

```{r}
ecodist::mantel(as.dist(cult.m2) ~
                as.dist(ling.m2),
                nperm = 10000)
```


Run a mantel test comparing the linguistic distances to the cultural distances, controlling for the historical distance between languages:

```{r}
ecodist::mantel(as.dist(cult.m2)~
                  as.dist(ling.m2) + 
                  as.dist(hist.m2),
                nperm = 10000)
```

# References

Bouckaert, Remco, Philippe Lemey, Michael Dunn, Simon J. Greenhill, Alexander V. Alekseyenko, Alexei J. Drummond, Russell D. Gray, Marc A. Suchard, and Quentin D. Atkinson (2012). Mapping the origins and expansion of the Indo-European language family. Science, 337(6097), 957-960.