---
title: "Kinship terms"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

```{r}
library(dplyr)
```

## Load data

```{r}
setwd("~/Documents/Bristol/word2vec/word2vec_DPLACE/processing/")

d = read.csv("../data/FAIR/nel-wiki-k100-alignments-merged-long.csv",stringsAsFactors = F)

```

Mean local alignment is high for kinship:

```{r}
x = d %>% group_by(SEMANTICFIELD)%>% summarise(
  local_alignment = mean(local_alignment)
) 
x$SEMANTICFIELD = factor(x$SEMANTICFIELD,levels=x$SEMANTICFIELD[order(x$local_alignment)])
ggplot(x,aes(y=SEMANTICFIELD,x=local_alignment)) +
  geom_point()
```

However, the "Kinship" category in the Concepticon includes many categories of word, including relational kin terms, kin groups, non-relational words for children, and pronouns:

```{r}
unique(d[d$l1=="en" & d$SEMANTICFIELD=="Kinship",]$Word_Form_l1)
```


If we plot the alignment by different type of words, it's clear that the kin terms are the most highly aligned. In fact, the mean alignment is higher than for Quantity.

```{r}
kinwords = c("Ehefrau::N","Bruder::N","Sohn::N","Ehemann::N",
             "Onkel::N","Eltern::N","Tochter::N","Vater::N",
             "Ehefrau::N","Mutter::N","Tante::N","Schwester::N",
             "Großvater::N","Großmutter::N")
pronounwords = c("ich::PRN", "er::PRN", "sie::PRN", "wir::PRN", "du::PRN", "ihr::PRN" )

d$kinwords = "Other"
d[d$Concept_ID %in% kinwords,]$kinwords = "Kin terms"
d[d$Concept_ID %in% pronounwords,]$kinwords = "Pronouns"

d$kinwords = factor(d$kinwords,levels=c("Kin terms","Pronouns","Other"))

ggplot(d[d$SEMANTICFIELD=="Kinship",], aes(x=kinwords,y=local_alignment)) +
  geom_violin() + geom_boxplot(width=0.3)

```

We can compare the alignment for the term "uncle" with the system of kin terms for uncles, from the Standard Cross-Cultural Sample:

```{r}
sccs = read.csv("../data/dplace_SCCS/data.csv",stringsAsFactors = F)
s = read.csv("../data/dplace_SCCS/societies.csv",stringsAsFactors = F)
sccs$glotto = s[match(sccs$soc_id,s$id),]$glottocode
sccs = sccs[sccs$var_id=="SCCS639",]
sccs = sccs[sccs$year>200,]

uncle = d[d$Concept_ID=="Onkel::N",]

v = read.csv("../data/dplace_SCCS/codes.csv",stringsAsFactors = F)
v = v[v$var_id=="SCCS639",]
sccs$uncleType = v[match(sccs$code,v$code),]$name

uncle$uncleType.l1 = sccs[match(uncle$Glottocode_l1,sccs$glotto),]$uncleType
uncle$uncleType.l2 = sccs[match(uncle$Glottocode_l2,sccs$glotto),]$uncleType

uncle = uncle[!is.na(uncle$uncleType.l1),]
uncle = uncle[!is.na(uncle$uncleType.l2),]

uncle$uncleType.cat = apply(uncle[,c("uncleType.l1","uncleType.l2"),],
                              1,function(X){paste(sort(X),collapse="/")})
uncle$uncleType.same = uncle$uncleType.l1==uncle$uncleType.l2
```

The data has `r nrow(uncle)` observations between `r length(unique(c(uncle$l1,uncle$l2)))` languages

```{r}
ggplot(uncle,aes(y=local_alignment,x=uncleType.cat)) + 
  geom_boxplot() + coord_flip()

ggplot(uncle,aes(y=local_alignment,x=uncleType.same)) + 
  geom_boxplot()
```

```{r}
uncle$sameFam = uncle$Language_Family_l1==uncle$Language_Family_l2
summary(lm(local_alignment ~ 1 + sameFam + uncleType.same, data = uncle))

```


```{r}
ea = read.csv("../data/dplace-data-1.0/csv/EA_data.csv",stringsAsFactors = F)
ea = ea[ea$VarID==12,]
v = read.csv("../data/dplace-data-1.0/csv/EACodeDescriptions.csv",stringsAsFactors = F)
v = v[v$VarID==12,]
ea$cat = v[match(ea$Code,v$Code),]$ShortName
ea$cat[ea$cat=="Missing data"] = NA

ea$cat2 = ea$cat
#ea$cat2 = factor(ea$cat2,levels=
#        c("Ambi-viri",  "Ambilocal","Neolocal", "Patrilocal","Separate","Virilocal"),
#        c("M",  "N","N", "M","N","M"))

g = read.csv("../data/dplace-data-1.0/csv/xd_id_to_language.csv",stringsAsFactors = F)
s = read.csv("../data/dplace-data-1.0/csv/ALL_soc_ids_to_xd_ids_24Feb2016.csv",stringsAsFactors = F)

ea$xid = s[match(ea$soc_id,s$soc_id),]$xd_id
ea$glotto = g[match(ea$xid,g$xd_id),]$DialectLanguageGlottocode

husband = d[d$Concept_ID=="Ehemann::N" | d$Concept_ID=="Ehefrau::N",]

ea = ea[ea$glotto %in% unique(c(uncle$Glottocode_l1,uncle$Glottocode_l2)),]
husband$locality.l1 = ea[match(husband$Glottocode_l1,ea$glotto),]$cat2
husband$locality.l2 = ea[match(husband$Glottocode_l2,ea$glotto),]$cat2

husband = husband[!is.na(husband$locality.l1),]
husband = husband[!is.na(husband$locality.l2),]

husband$locality.same = husband$locality.l1 == husband$locality.l2

ggplot(husband,aes(x=locality.same,y=local_alignment)) +
  geom_boxplot()
husband$sameFam = husband$Language_Family_l1==husband$Language_Family_l2
summary(lm(local_alignment ~ 1 + sameFam + locality.same + Concept_ID, data = husband))

```

